\begin{acknowledgementsen}

This dissertation would not have been possible without the help of many people along the way, to whom I owe my deepest gratitude.

First of all, I would like to thank Professor Jane Hsu and Professor Daniel Shih, my two supervisors at NTU, for all their time and support. Professor K. J. Lin, for starting the project that gave me the opportunity to work at the Intel-NTU Connected Context Computing Center, and that indeed eventually led to this work. And Professor Pai H. Chou, for always making the time to have an inspiring talk and for your valuable feedback on my writings.

Special thanks to Professor Koen Langendoen, who, as my first supervisor during my time at Delft University, taught me most of what I know about doing research. I still use the things you taught me on an almost daily basis, both inside and outside of research. 

This dissertation builds on earlier work by Joshua Ellul. Thanks for all the discussions we had over the years and for coming to Taipei for what turned out to be a very nice collaboration.

During the seven years it took to complete this dissertation, I had the privilege to be able to combine my research in Taiwan with work at DSW Zorgverzekeraar in The Netherlands. My thanks go out to Jeroen de Haan and Frans ten Brink for making this possible, and to my colleagues, in particular Edwin Haddeman and Jeroen Staphorsius for taking an interest in my work and for your useful comments.

Living in a foreign country is immensely rewarding, but not without challenges. From the beginning of my stay in Taiwan, Bormann Chen has helped me out on countless occasions. Thanks for being such a good friend, and for still not giving up on me learning Chinese.

I am grateful for the support of many friends and family, but a few deserve special mention. Thank you Gwena, for all the late night drinks in your living room, inspiring conversations, and exploring many interesting places in and around Taipei together. The long process of completing a PhD can be lonely at times. Three people in particular often kept me company while we were each working on our own projects, this has been a great help over the last few years. Thank you Ami, for spending many days studying together, and for always taking care of my cat while I'm away. Dayton, for being my partner in discovery of many new coffee shops in Taipei in which to work, for all the discussions we had about our theses - it was great to share experiences with someone going through the same process - and for the many games of chess to relax after a good day's work. And Ching-chi, for all your support, for keeping me company during many late nights in the lab, and for all the after work runs.

And finally thanks to my parents, my first and most important teachers.
  
\vspace{2cm} 

This research was supported in part by the Ministry of Science and Technology of Taiwan (MOST 105-2633-E-002-001), National Taiwan University (NTU-105R104045), Intel Corporation, and Delta Electronics.
\end{acknowledgementsen}
